---
title: "Tutorial on the Least Squares Rating System"
output:
  pdf_document: default
  html_notebook: default
urlcolor: blue
linkcolor: red
---

```{r, message=F, warning=F}
library(data.table)
```

This R notebook is an exercise and an illustration of the mathematical [paper](https://www.tandfonline.com/doi/abs/10.1080/09332480.1995.11884566) of [Hal Stern](https://www.ics.uci.edu/~sternh/) at the journal [Chance](https://chance.amstat.org/). His paper is entitled "Who's Number 1 in College Football? ... And How Might We Decide?"

# Load and inspect the data

In this tutorial we are going to implement various algorithms related to the Least Squares Rating System. The data we need is as follows

  * the schedule of season 1994.
  * the ranking and ratings of the 107 Division 1-A teams.


The data with the match schedule of season 1994.

```{r}
x = fread("./data/1994-season.csv")
x
```

The ranking and ratings of the 107 Division 1-A teams

```{r}
r = fread("./data/1994-rankings.csv")
r
```


Let us separate the top teams from the lower-divisional teams.

```{r}
teams = sort(unique(c(x$Winner, x$Loser)))
top_teams = sort(unique(r$School))
low_teams = setdiff(teams, top_teams)
```


# Examples 

## Find the normal equations of the Least Squares Rating System with home advantage set to 3 points.

The solution is provided in the source file


```{r}
source("lsq-functions.R")
```

```{r}
l = lsq.system.ha(x, 3)
l$game[1:10, 1:10]
```
## Solve the linear system and rank-order the teams according to rating

Define the coefficients and parameters
```{r}
game = l$game; pd = l$pd; d = nrow(game)
```


Add a linear constraint to the system. 

```{r}
#all ratings sum to 0
game[d, ] = rep(1,d) 
pd[d] = 0
#solve the linear system
sol = solve(game,pd)
```


Let us now rank-order the teams and display their ratings.

```{r}
rorder = order(rank(sol[1:d]), decreasing = T)
rankings = data.table(team = teams[rorder], rating = round(sol[rorder],2))
rankings[1:10]
```

## Write a function that implements all of the above steps

All these steps are implemented in `lsq.ratings.ha`.

```{r}
lsq.ratings.ha(x, 3)[1:10]
```

## Implement the Jackobi iteration method to solve a linear system and compare it with the built-in solver

We have already implemented the Jacobi iteration as an example, which we load from source file.

```{r}
source("iter-methods.R")
```

We run the function `Jacobi`. It prints the number of iterations it makes in order to achieve the required tolerance level.

```{r}
sol.ja = Jackobi(game, pd)
```
In this case, the function made 99 iterations. Let us see the first five solutions

```{r}
sol.ja[1:5]
```
Let us compare them with the previous solution.

```{r}
sol[1:5]
```

# Problems about the LSqRatings

## Modify the above rating system so that the mean rating of all lower-divisional teams is zero 

## Modify the above rating system so that the rating of a particular team is zero 

## Compute the LSqRatings and the home advantage from the data

## Compute the LSqRatings and the home advantage from the data with all lower-divisional teams having rating exactly 0

## Decompose the ratings in offence and defence parts

## Compute the LSqRatings and the home advantage from the data with given weights for each match

# Problems about iterative numerical computation of linear systems

## Implement the Jackobi algorithm and solve some of the above systems

## Implement the Gauss-Seidel algorithm and solve some of the above systems





